Getting Started with the Graded Response Model (GRM): A gentle introduction and tutorial in R

### 2024-01-22 manuscript.qmd
creating blind annotated R script and revising manuscript

---
title: "Getting Started with the Graded Response Model (GRM): A gentle introduction and tutorial in R"
shorttitle: "A Graded Response Model Tutorial"
author:
  - name: Rizqy Amelia Zein
    corresponding: true
    orcid: 0000-0001-7840-0299
    email: amelia.zein@psikologi.unair.ac.id
    affiliations:
      - name: Universitas Airlangga
        department: Department of Psychology
        address: Jalan Airlangga 4-6
        city: Surabaya
        region: Jawa Timur, Indonesia
        postal-code: "60286"
      - name: Ludwig-Maximilians-Universität
        department: Department of Psychology
        address: Leopoldstraße 13
        city: Munich
        postal-code: "80802"
  - name: Hanif Akhtar
    orcid: 0000-0002-1388-7347
    affiliations:
      - name: ELTE Eötvös Loránd University
        department: Doctoral School of Psychology
        country: Hungary
      - name: Universitas Muhammadiyah Malang
        department: Faculty of Psychology
        city: Malang
        region: Jawa Timur, Indonesia
abstract: "This tutorial paper introduces the Graded Response Model (GRM), a tool for testing measurement precision under the Item Response Theory (IRT) paradigm. Addressing common problems of measurement imprecision and lack of construct validity, the tutorial guides researchers through a unidimensional GRM analysis in the R environment, using psych, mirt, and ggmirt packages. GRM is specifically designed to examine the psychometric properties of psychological scales with polytomous (Likert-style) items. The tutorial illustrates the procedure using data from the Open Psychometrics Database on the Right-Wing Authoritarianism (RWA) scale, outlining the theoretical underpinnings of GRM, and steps for data preparation, model fitting, interpretation of results, and dealing with some common issues that may typically arise in the process."
keywords:
  - graded response model
  - item response theory
  - mirt
  - ggmirt
  - psych
  - R
author-note:
  disclosures:
    conflict-of-interest: "We declare we have no competing interests."
    financial-support: "The first author receives a PhD scholarship from the German Academic Exchange Service (Deutscher Akademischer Austauschdienst, DAAD)."
lang: en
bibliography: references.bib
format:
  apaquarto-docx: default
editor: visual
execute: 
  warning: false
  error: false
  output: false
  echo: true
---

@@

The foundation of scientific research is the measurement process, which involves defining the construct to be measured, determining measurement assumptions, and evaluating measurement validity and precision. To illustrate this process, consider a researcher studying the role of authoritarian personality in explaining prejudice must first clearly define these variables, and then, develop a measurement strategy to operationalize the variables. This is challenging for applied psychology researchers, as most psychological constructs are not directly observable. To ensure a precise measurement, researchers rely on statistical techniques, assuming that these non-observable traits are "out there," or later we name this concept a *latent trait* or a *manifest variable*. Researchers then create tools, like scales, stimuli, tasks, observation checklists, etc., which are assumed to represent these latent traits.

In psychological research, where variables are often not directly observable, construct validity, which refers to a question on whether a scale measures what it is intended to measure [@cronbachConstructValidityPsychological1955], becomes crucial. Construct validity can be scrutinized by linking test scores to other related constructs and observations [@cronbachConstructValidityPsychological1955]. Test validity relies on whether the construct behaves as theory predicts within a "nomological network" [@cronbachConstructValidityPsychological1955]. In practice, researchers establish validity by correlating the test score with external criterion i.e., other theoretically related traits or behavior [@zumboSettingStageValidity2014].

Validity theory has evolved and has been expanded from time to time and validity is now seen not as a property of a specific test, but rather lies in the consequential aspects of the test use [@kaneValidatingInterpretationsUses2013; @kaneValidityEvaluationClaims2016; @messickValidityPsychologicalAssessment1995]. In this sense, a test is deemed valid when researchers can demonstrate evidence for formulating a logical argument [@kaneValidatingInterpretationsUses2013; @kaneValidityEvaluationClaims2016], such as causal inference [@rutkowskiCausalInferencesLarge2016], used to defend test interpretations and its practical use [@messickValidityPsychologicalAssessment1995]. This makes construct validity an ongoing process where researchers are advised to continuously gather and report evidence for validity of the measures they use in their research [@flakeConstructValidationSocial2017; @messickValidityPsychologicalAssessment1995; @zumboSettingStageValidity2014].

Yet in practice, researchers tend to overlook to the quality of their measures. A study reviewing articles published in a social and personality psychology journal found that only a little more than half (53%) of sample articles cited validity evidence from previous validation studies. Nineteen percent of those articles, nonetheless, had adapted or modified the measures in many different ways so that the evidence for validity may not extend to these modified versions [@flakeConstructValidationSocial2017]. Additionally, a larger study examining fifteen commonly used measures in social and personality psychology (e.g., Big Five Inventory, belief in just world scale, need for cognitive closure scale, etc.) even paints a grimmer picture [@husseyHiddenInvalidity152020]. While the majority of the tested scales (88%) purportedly had good validity, a more exhaustive examination through internal consistency, test-retest reliability, factor structure, and invariance tests revealed that only 4% of these scales possess good measurement validity [@husseyHiddenInvalidity152020]. Furthermore, applied psychological researchers seem to overlook the assumptions underlying their chosen method for evaluating reliability. For example, while Cronbach's $\alpha$ is always almost reported [@flakeStrengtheningFoundationEducational2021] as the most popular reliability coefficient [@mcneishThanksCoefficientAlpha2018], it is unclear whether researchers are aware of its underlying assumptions (that true scores and item variances should be equal and that item residuals should be uncorrelated, i.e., $\tau$ equivalence), which limits its usefulness [@sijtsmaUseMisuseVery2008].

In this sense, measurements oversights can raise legitimate concerns about the replicability of research findings because measurement directly contributes to the (in)validity of the findings [@flakeMeasurementSchmeasurementQuestionable2020; @lilienfeldPsychologicalMeasurementReplication2020]. To better identify how these oversights manifest themselves in the research process, we should pay more attention to decisions that researchers make that potentially compromise the validity of the measures used in their studies, a phenomenon known as "*questionable measurement practices*" [@flakeMeasurementSchmeasurementQuestionable2020]. Furthermore, the lack of evidence for measurement validity contributes to the difficulty of reaching a consensus on the best way to measure a construct. In the long run, this is a major obstacle to the establishment of cumulative psychological science [@elsonPsychologicalMeasuresAren2023]. Given these circumstances, it is critical to provide applied psychological researchers with a set of tools and practical steps to help them assess measurement validity so that they can improve their measurement practices, and thus, the credibility of their findings.

The purpose of this article is to provide a brief introduction to the Graded Response Model (GRM), a family of item response models specifically designed to assess the measurement precision of a polytomous (Likert-style) scale. We begin this paper by briefly explaining the basics of Item Response Theory (IRT), the theoretical foundation of GRM, and contrasting its differences with the widely known Classical True Score Theory (CTT). In this paper, we aim to keep our tutorial concise, so we do not cover all IRT concepts in detail. Interested readers are referred to available didactic texts on IRT, such as @embretsonItemResponseTheory2000 for the introductory level, and @bakerBasicsItemResponse2017 and @deayalaTheoryPracticeItem2022 for a more technical and comprehensive overview.

Further, we briefly explain what GRM is and how a GRM model is specified. In this part, we highlight how model and item parameters are specified and interpreted as well as visualizations (i.e., plots) derived from a GRM analysis. We also introduce three key assumptions underlying a GRM and how to test these assumptions. We then provide an illustrative example of testing measurement precision using GRM, and in this part, we show how to practically implement GRM analysis and interpret its results. We complement the illustrative example with code in the open-source programming language R [@rcoreteamLanguageEnvironmentStatistical2023] so that the readers can implement and reproduce the example presented in this paper. Readers with some experience in R and RStudio can easily follow this tutorial, but for those who are not yet familiar with R, we refer them to excellent, easy-to-follow materials developed by @navarroPsychologicalScience2018. At last, we present some concluding remarks to encourage researchers to implement a GRM analysis as a part of their routine.

@@

To scrutinize their measurement quality, psychological researchers have relied upon two foundational principles underlying psychological testing, i.e., CTT and IRT, to decompose observed score into its *deterministic* and *random* or stochastic elements [@zumboValidityFoundationalIssues2006]. On the one hand, CTT decomposes observed scores into *true* score and its *error* components. In practice, researchers implement this principle by conducting Structural Equation Modeling (SEM), Exploratory Factor Analysis (EFA), or Confirmatory Factor Analysis (CFA). Specifying the measurement part of SEM, or employing EFA and CFA models, allows for the decomposition of the covariance matrix of a given data into latent factors, which represent true score or trait level, and error terms (i.e., residuals or unique variances). Consequently, the relationship between observed score and true score is *linear*, since true score is essentially a linear transformation of observed score [@embretsonItemResponseTheory2000]. IRT, on the other hand, models this relationship in a *probabilistic* rather than a linear fashion. More sharply, an IRT model assumes a probabilistic relationship between the observed score and the latent trait being measured ($\theta$), and in particular, this probability takes into account the properties of the items, such as item discrimination, difficulty, and, when appropriate, guessing [@bakerBasicsItemResponse2017].

The most critical distinction between CTT and IRT is how they conceptualize measurement error. In CTT, standard deviation of the observed score (*s*) is used to calculate standard error of measurement. Since measurement error relies on *s*, CTT assumes that measurement error is sample-dependent and constant for all individuals in the sample regardless of their trait level. This shortcoming limits the utility of CTT, especially since psychologists often need to interpret individual scores, not just evaluate the test as a whole. Additionally, since standard error of measurement is also a function of reliability, which is typically denoted as a correlation between two parallel tests, reliability is assumed to be the same across different levels of a trait or ability. This is unrealistic since researchers often encounter situations where a test is overly difficult for a group of low-performing participants, providing little information beyond an indication that their trait or ability level is significantly lower than what the test can measure.

IRT offers an elegant solution to this issue by enabling the calculation of standard errors for each individual (*SE*~$\theta$~), and test reliability can be inferred from the average of *SE*~$\theta$~ across a sample of participants [@embretsonItemResponseTheory2000; @langSciencePracticeItem2021]. This approach allows for the estimation of reliability at varying levels of trait or ability (i.e., test information function - TIF). By doing this, when a test is overly challenging for low-performing individuals or too easy for high-performing individuals, IRT can help identify the levels of trait or ability at which the test is most reliable. A concrete example of this is a study that examined the reliability of the Short Dark Tetrad (SD4) scale from an IRT perspective [@blotnerDarkTriadDead2022]. According to this study, the sadism subscale of the SD4 scale is the most reliable for measuring individuals with average to high levels ($\bar{x}$ \< $\theta$ \< 2.5 *SD*) of sadism but suboptimal for measuring individuals with low levels of sadism [@blotnerDarkTriadDead2022]. Therefore, IRT models offer a more informative and nuanced perspective that proves useful for researchers who wish to closely examine the performance of their measures.

Further, IRT allows researchers to examine the performance of a specific item (hence "item" response theory) by specifying the relationship between item score and $\theta$ given one-, two-, or three-parameter. The one-parameter logistic (1PL) or Rasch model, the simplest variant of the IRT models, accounts only item difficulty (*b*) while assuming equal item discrimination (*a*) across items. In practice, this model overlaps slightly with CTT in the sense that they assume that all items have equal informative value, so that the estimated $\theta$ of a 1PL/Rasch model is identical to the sum score of CTT [@langSciencePracticeItem2021; @stemlerRaschMeasurementItem2021]. This notion actually stems from a philosophical principle of *specific objectivity*, which requires that comparisons between measures remain independent of both the item and individual characteristics [@raschSpecificObjectivityAttempt1977]. Put simply, Rasch model assumes that the latent trait ($\theta$) should remain unaffected by specific items used in the test. For example, the difference in depression levels between two individuals should always be the same regardless of the scale used to measure their depression levels. While this principle reasonably enforces objectivity in measurement practices, it requires a strong theory delineating the construct and strict requirements of data-model fit while both of these assumptions rarely hold in a real-world scenario, especially in psychological science.

Furthermore, the two-parameter logistic (2PL) model fills this gap by allowing for differences in item ability to differentiate individuals with varying levels of $\theta$ (i.e., item discrimination parameter - *a*). Additionally, in some contexts, researchers may suspect that some parts of the probabilistic relationship between observed score and $\theta$ is explained by guessing, thus three-parameter logistic (3PL) model incorporates guessing parameter (*c*). The field of IRT has rapidly evolved to include various models suited to specific contexts, such as handling ordinal responses [@murakiGeneralizedPartialCredit1992; @samejimaGradedResponseModel1997], categorical responses [@thissenNominalCategoriesItem2013], or assessing multidimensional traits simultaneously [@bockMarginalMaximumLikelihood1981; @chalmersMirtMultidimensionalItem2012]. It is important to note that 1PL, 2PL, and 3PL models are only applicable to binary or dichotomous data (e.g., true/false responses), and the focus of this article is nonetheless to demonstrate the utility of IRT for fitting the ordered (Likert-style) responses. We briefly summarize the features of the most commonly used IRT models in {apatb-table1}.

@@

The GRM is a family of IRT models specifically designed to analyze ordered polytomous (Likert-style) data [@samejimaGradedResponseModel1997; @samejimaGeneralGradedResponse2010; @samejimaGradedResponseModels2016]. The fundamental idea of GRM is to extend the logic of simpler dichotomous IRT models to polytomous scales by applying a probabilistic function to each response category. In dichotomous 2PL IRT models, the probability of answering a correct response is modeled as a function of $\theta$, item difficulty, and item discrimination. GRM extends this idea by modeling the probability of picking a certain response category or higher on an item (i.e., *step function*) given to item difficulty and discrimination. For each response category (e.g., answering "agree" on a five-point Likert scale), GRM calculates the cumulative probability of a participant answering "agree" or above, given to their $\theta$. Therefore, the notion of item difficulty in dichotomous IRT models is extended to the step function (i.e., item threshold) to handle ordered polytomous data.

@@

Let's take a closer look at {apafg-fig1}, which shows an example of an item probability function (IPF) from an item on a scale with five response categories. In general, IPF describes a relationship between $\theta$ (x-axis) and the probability of endorsing a response category (y-axis - *P*$\theta$). {apafg-fig1} consists of five category probability curves, each of which represents the probability of endorsing a response category given $\theta$. Item threshold is exactly the location of $\theta$ where two adjacent category probability curves cross each other.

@@

Before running an analysis, researchers need to be aware of the assumption underlying a GRM model. The first key assumption is that $\theta$ is monotonically correlated with the probability of endorsing a response category. In other words, individuals with higher levels of $\theta$ have a higher probability of endorsing response category that represents greater intensity of the traits [@hambletonIRTModelsAnalysis2010]. For instance, people with a higher tendency of sadistic personality have a greater chance of answering "*moderately agree*" or "*strongly agree*" on a sadism scale compared to those with lower levels of sadism.

@@

Relatedly, a GRM model assumes that the item pool is locally independent, as the third assumption. Local independence suggests that participants' responses to one item do not influence their response to another. It is important to note that evidence for unidimensionality does not always warrant local independence. As an eyeball example - consider trust in science scale containing these two items: "*I trust scientists working in natural science*" and "*Most biologists are trustworthy*." While both items might measure the same latent trait (i.e., trust in science), those who trust scientists working in natural science are very likely to also trust biologists. To inspect whether items are locally independent, researchers can examine the relationships between item responses after accounting for $\theta$ using residual correlation analysis [@chenLocalDependenceIndexes1997].

@@

In this paper, we use the latest version of the RWA scale [@altemeyerAuthoritarians2006], which consists of 22 items that participants are asked to respond with their agreement to those items on a nine-point scale, ranging from "*strongly disagree*" (-4) to "*strongly agree*" (+4). While the RWA scale consists of three sub-dimensions, i.e., submissiveness, aggression, and conventionalism, the RWA scale is (theoretically) unidimensional because [@altemeyerAuthoritarians2006] shows that the sub-dimensions are strongly intercorrelated. While some evidence shows otherwise [@duckittMultidimensionalityRightWingAuthoritarian2013], for the purpose of this article, we assume that the RWA scale is unidimensional.

@@

To maximize the reproducibility of our illustrative example, we wrote the article as a Quarto (.qmd) document, where we integrate the R codes used in the analysis as well as its outputs. We also include the annotated R script file (.R) for the example we used as a supplementary document, which we highly recommend to use for a didactic purpose. The Quarto file (and its corresponding .docx and .pdf output) and R script file are publicly available on [a Github repository](https://github.com/rameliaz/grm-tutorial-paper). The dataset we used in this tutorial paper was obtained from [Open-Source Psychometrics Project](http://openpsychometrics.org/_rawdata/), which is publicly accessible.

@@

This command summarizes key descriptive statistics of a data frame, and the summary is provided in {apatb-table2}. As we see in {apatb-table2}, all variables have a minimum value of `0`, while the minimum score should be `1`. Again, there is no information provided in the codebook regarding what `0` actually represents. Therefore, before deciding what we should do next, it is better to find out how many `0` we have in the data frame. To this end, we can calculate the percentage of `0` in each item by running these commands:

@@

After executing these commands, readers can view the percentage of `0` in each items in their R console, which overall are very small, ranging from 0.09% to 0.32% of the total cases. Since the percentages are very small, `0` is likely a code for a missing value. It is important to note that *mirt* package uses the Expected-Maximization (EM) algorithm to estimate model parameters, and the EM algorithm can handle missing values [@chalmersMirtMultidimensionalItem2012]. If we decide to eliminate cases with missing values, we still have a sizable number of cases, nonetheless. Therefore, to simplify our analysis, we only keep cases with complete responses by executing these commands:

@@

Interestingly, if we look at {apatb-table2} again, the mean scores of the items seem to differ drastically. Some items have high mean scores, ranging from 6 to 7, but some of those are comparatively low. The RWA scale indeed has some unfavorable items but, again, there is no information in the codebook whether these unfavorable items have been reversely coded. However, if we look closely at ten items with high mean ($\bar{x}$ > 6), these are the items that are supposed to be unfavorable. Therefore, we can reasonably assume that unfavorable items have not been reversed in the original dataset, and we should do so by running these commands:

@@

In this step, we start to specify a vector representing our model specification, such as:

@@

```{r}
M2(fit, type = "C2")
```
```{r}
#| include: false

>>>>

---
title: "Getting Started with the Graded Response Model (GRM): A gentle introduction and tutorial in R"
shorttitle: "The Graded Response Model Tutorial"
author:
  - name: Rizqy Amelia Zein
    corresponding: true
    orcid: 0000-0001-7840-0299
    email: amelia.zein@psikologi.unair.ac.id
    affiliations:
      - name: Universitas Airlangga
        department: Department of Psychology
        address: Jalan Airlangga 4-6
        city: Surabaya
        region: Jawa Timur, Indonesia
        postal-code: "60286"
      - name: Ludwig-Maximilians-Universität
        department: Department of Psychology
        address: Leopoldstraße 13
        city: Munich
        postal-code: "80802"
  - name: Hanif Akhtar
    orcid: 0000-0002-1388-7347
    affiliations:
      - name: ELTE Eötvös Loránd University
        department: Doctoral School of Psychology
        country: Hungary
      - name: Universitas Muhammadiyah Malang
        department: Faculty of Psychology
        city: Malang
        region: Jawa Timur, Indonesia
abstract: "This tutorial introduces the Graded Response Model (GRM), a tool for testing measurement precision within the Item Response Theory (IRT) paradigm. Addressing common problems of measurement imprecision and lack of construct validity, the tutorial guides researchers through a unidimensional GRM analysis in the R environment, using psych, mirt, and ggmirt packages. GRM is specifically designed to examine the psychometric properties of psychological scales with polytomous (Likert-style) items. The tutorial illustrates the procedure using data from the Open Psychometrics Database on the Right-Wing Authoritarianism (RWA) scale, outlining the theoretical underpinnings of GRM, and steps for data preparation, testing model assumptions, model fitting, and interpretation of results."
keywords:
  - graded response model
  - item response theory
  - mirt
  - ggmirt
  - psych
  - R
author-note:
  disclosures:
    conflict-of-interest: "We declare we have no competing interests."
    financial-support: "The first author receives a PhD scholarship from the German Academic Exchange Service (Deutscher Akademischer Austauschdienst, DAAD)."
lang: en
bibliography: references.bib
format:
  apaquarto-docx: default
editor: visual
execute: 
  warning: false
  error: false
  output: false
  echo: true
---

@@

Only half of published findings include evidence of measurement validity [@flakeConstructValidationSocial2017] and a large proportion of published studies suffer from hidden invalidity [@husseyHiddenInvalidity152020]. Lack of evidence for measurement validity can raise serious concerns about the replicability of research findings, as it directly contributes to the invalidity of the findings [@flakeMeasurementSchmeasurementQuestionable2020; @lilienfeldPsychologicalMeasurementReplication2020]. Furthermore, the lack of evidence for measurement validity can hamper consensus on the best way to measure a construct, which becomes a major obstacle to the establishment of cumulative psychological science [@elsonPsychologicalMeasuresAren2023]. Given these circumstances, it is critical to provide applied psychological researchers with a set of tools and practical steps to help them assess measurement validity so that they can ensure the credibility of their findings.

The purpose of this article is to provide a brief introduction to the Graded Response Model (GRM), a family of item response models specifically designed to assess the measurement precision of a polytomous (Likert-style) scale. We begin this paper by briefly explaining the basics of Item Response Theory (IRT) and contrasting its differences with the widely known Classical True Score Theory (CTT). In this paper, we aim to keep our tutorial concise, so we do not cover all IRT concepts in detail. Interested readers are referred to available didactic texts on IRT, such as @embretsonItemResponseTheory2000 for the introductory level, and @bakerBasicsItemResponse2017 and @deayalaTheoryPracticeItem2022 for a more technical and comprehensive overview.

Further, we briefly explain what GRM is and how it is specified. In this part, we highlight how model and item parameters are specified and interpreted as well as visualizations derived from a GRM analysis. We also introduce three key assumptions underlying a GRM and how to test these assumptions. We then provide an illustrative example of testing measurement precision using a GRM analysis, and in this part, we show how to practically implement a GRM analysis and interpret its results. We complement the illustrative example with code in the open-source programming language R [@rcoreteamLanguageEnvironmentStatistical2023] so that the readers can implement and reproduce the example presented in this paper. Readers with some experience in R and RStudio can easily follow this tutorial, but for those who are not yet familiar with R, we refer them to excellent, easy-to-follow materials developed by @navarroPsychologicalScience2018. Finally, we present some concluding remarks to encourage researchers to implement a GRM analysis as a part of their routine.

@@

To scrutinize their measurement quality, psychological researchers have relied on two basic principles underlying psychological testing, i.e., CTT and IRT, to decompose observed scores into their *deterministic* and *random* or stochastic components [@zumboValidityFoundationalIssues2006]. CTT, on the one hand, decomposes observed outcomes into *true* scores and *error* components. In practice, researchers implement this principle by performing structural equation modeling (SEM), exploratory factor analysis (EFA), or confirmatory factor analysis (CFA). Specifying the measurement part of a SEM model, or estimating EFA and CFA models, allows the covariance matrix of a given set of data to be decomposed into latent factors, which represent the true score or trait level, and error terms (i.e., residuals or unique variances). Consequently, the relationship between observed score and true score is *linear*, since true score is essentially a linear transformation of observed score [@embretsonItemResponseTheory2000]. IRT, on the other hand, models this relationship in a *probabilistic* rather than a linear fashion. More specifically, an IRT model assumes a probabilistic relationship between the observed score and the latent trait being measured ($\theta$), and in particular, this probability takes into account item properties such as item discrimination, difficulty, and, when appropriate, guessing [@bakerBasicsItemResponse2017].

The most critical distinction between CTT and IRT is how they conceptualize measurement error. In CTT, standard deviation of the observed score (*s*) is used to calculate standard error of measurement. Therefore, CTT assumes that measurement error is sample-dependent and constant for all individuals in the sample regardless of their trait level. This shortcoming limits the utility of CTT, especially since psychologists often need to interpret individual scores, rather than just evaluate the test as a whole. Additionally, since standard error of measurement is also a function of reliability, reliability is assumed to be the same across different levels of a trait or ability. This assumption is unrealistic since researchers often encounter situations where a test is overly difficult for a group of low-performing participants, providing little information beyond an indication that their trait or ability level is significantly lower than what the test can measure.

IRT offers an elegant solution to this issue by allowing the calculation of standard errors for each individual (*SE*$\theta$), and thus, test reliability can be inferred from the average of *SE*$\theta$ across a sample of participants [@embretsonItemResponseTheory2000; @langSciencePracticeItem2021]. This approach allows for the estimation of reliability at varying levels of trait or ability (i.e., test information function - TIF). By doing this, when a test is overly challenging for low-performing individuals or too easy for high-performing individuals, IRT can help identify the levels of trait or ability at which the test is most reliable. A concrete example of this is a study that examined the reliability of the Short Dark Tetrad (SD4) scale from an IRT perspective [@blotnerDarkTriadDead2022]. According to this study, the sadism sub-scale of the SD4 is the most reliable for measuring individuals with average to high levels ($\bar{x}$ \< $\theta$ \< 2.5 *SD*) of sadism but suboptimal for measuring individuals with lower levels of sadism [@blotnerDarkTriadDead2022]. In this sense, IRT models offer a more informative and nuanced perspective that proves useful for researchers who wish to closely examine the performance of their measures.

Further, IRT allows researchers to examine the performance of a specific item by specifying the relationship between item score and $\theta$ given one-, two-, or three-parameter. The one-parameter logistic (1PL) or a Rasch model, the simplest variant of the IRT models, accounts for only item difficulty (*b*) while assuming equal item discrimination (*a*) across items. In practice, this model overlaps slightly with CTT in the sense that they assume that all items have equal informative value, so that the estimated $\theta$ of a 1PL/Rasch model is identical to the sum score of CTT [@langSciencePracticeItem2021; @stemlerRaschMeasurementItem2021]. This notion actually stems from a philosophical principle of *specific objectivity*, which requires that comparisons between measures remain independent of both the item and individual characteristics [@raschSpecificObjectivityAttempt1977]. Put simply, a Rasch model assumes that the latent trait ($\theta$) should remain unaffected by specific items used in the test. More concretely, the difference in depression levels between two individuals should always be the same regardless of the scale used to measure their depression levels. While this principle reasonably enforces objectivity in measurement practices, it requires a strong theory underlying the construct and strict requirements of data-model fit, and both of these assumptions rarely hold in a real-world scenario, especially in psychological science.

Furthermore, the two-parameter logistic (2PL) model fills this gap by allowing for differences in item ability to differentiate individuals with varying levels of $\theta$ (i.e., item discrimination parameter - *a*). In addition, in some contexts, researchers may suspect that some parts of the probabilistic relationship between observed score and $\theta$ is explained by guessing, thus three-parameter logistic (3PL) model incorporates guessing parameter (*c*). The field of IRT has rapidly evolved to include various models suited to specific contexts, such as handling ordinal responses [@murakiGeneralizedPartialCredit1992; @samejimaGradedResponseModel1997], categorical responses [@thissenNominalCategoriesItem2013], or assessing multidimensional traits simultaneously [@bockMarginalMaximumLikelihood1981; @chalmersMirtMultidimensionalItem2012]. It is important to note that 1PL, 2PL, and 3PL models are only applicable to binary or dichotomous data (e.g., true/false responses), and the focus of this article is nonetheless to demonstrate the utility of IRT for fitting the ordered (Likert-style) responses. We briefly summarize the features of the most commonly used IRT models in {apatb-table1}.

```{r}
#| label: apatb-table1
#| apa-cap: Comparison Between The Commonly Used IRT Models
#| apa-note: NULL
#| echo: false
```

+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Model                                   | Key Characteristics                                                                                                                                         | Data Type   | Response Options        |
+=========================================+=============================================================================================================================================================+=============+=========================+
| 1-PL Model (Rasch Model)                | 1\. Estimates only item difficulties (*b*).                                                                                                                 | Dichotomous | Correct/Incorrect (0/1) |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Assumes that all items have equal discrimination parameters (*a*).                                                                                      |             |                         |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 3\. Item and person parameters are independent.                                                                                                             |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| 2-PL Model                              | 1\. Estimates item difficulties (*b*) and item discrimination (*a*).                                                                                        | Dichotomous | Correct/Incorrect (0/1) |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Less stringent than 1-PL model since it allows item discrimination parameters (*a*) to vary.                                                            |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| 3-PL Model                              | 1\. Estimates item difficulties (*b*), discrimination (*a*), and pseudo-guessing parameter (*c*).                                                           | Dichotomous | Correct/Incorrect (0/1) |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Appropriate for modeling a test data with multiple responses (e.g., multiple-choice tests), and thus, guessing might influence participants' responses. |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Graded Response Model (GRM)             | 1\. Appropriate for modeling ordinal data with more than two response categories (i.e., Likert-style).                                                      | Polytomous  | Ordered Categories      |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Estimates a discrimination parameter (*a*) and multiple threshold parameters (*b*) per item.                                                            |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Partial Credit Model (PCM)              | 1\. An extension of the 1-PL (Rasch) model for polytomous items.                                                                                            | Polytomous  | Ordered Categories      |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Estimates thresholds between adjacent categories but assumes equal discrimination across items.                                                         |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Generalized Partial Credit Model (GPCM) | Extending PCM to allow differential discrimination parameters across items.                                                                                 | Polytomous  | Ordered Categories      |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+
| Nominal Response Model (NRM)            | 1\. Appropriate for modeling categorical responses with no order.                                                                                           | Categorical | Unordered Categories    |
|                                         |                                                                                                                                                             |             |                         |
|                                         | 2\. Estimates discrimination parameters (*a*) and multiple category-specific parameters (*b*).                                                              |             |                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+-------------------------+

@@

The GRM is a family of IRT models specifically designed to analyze ordered polytomous (Likert-style) data [@samejimaGradedResponseModel1997; @samejimaGeneralGradedResponse2010; @samejimaGradedResponseModels2016]. The fundamental idea of GRM is to extend the logic of simpler dichotomous IRT models to polytomous scales by applying a probabilistic function to each response category. In dichotomous 2PL IRT models, the probability of answering a correct response is modeled as a function of $\theta$, item difficulty (*b*), and item discrimination (*a*). GRM extends this idea by modeling the probability of selecting certain response category or higher on an item (i.e., *step function*) given to item difficulty (*b*) and discrimination (*a*). For each response category (e.g., answering "agree" on a five-point Likert scale), GRM calculates the cumulative probability of a participant answering "agree" or above, given to their $\theta$. Therefore, the notion of item difficulty in dichotomous IRT models is extended to the step function (i.e., item threshold - *b*) to handle ordered polytomous data.

@@

{apafg-fig1} shows an example of an item probability function (IPF) from an item on a scale with five response categories. In general, IPF describes a relationship between $\theta$ (x-axis) and the probability of endorsing a response category (y-axis - *P*$\theta$). {apafg-fig1} consists of five category probability curves, each of which represents the probability of endorsing a response category given $\theta$. Item threshold is exactly the location of $\theta$ where two adjacent category probability curves cross each other.

@@

The first key assumption of a GRM is that $\theta$ is monotonically correlated with the probability of endorsing a response category. More sharply, individuals with higher levels of $\theta$ should have a higher probability of endorsing response category that represents greater intensity of the traits [@hambletonIRTModelsAnalysis2010]. For instance, people with a higher tendency of sadistic personality have a greater chance of answering "*moderately agree*" or "*strongly agree*" on a sadism scale compared to those with lower levels of sadism.

@@

Relatedly, a GRM model assumes that the item pool is locally independent, as the third assumption. Local independence suggests that participants' responses to one item do not influence their response to another. It is important to note that evidence for unidimensionality does not always warrant local independence. As an eyeball example - consider a scale measuring trust in science, which contains: "*I trust scientists working in natural science*" and "*Most biologists are trustworthy*." While both items might measure the same latent trait (i.e., trust in science), those who trust scientists working in natural science are very likely to also trust biologists. To test whether items are locally independent, researchers can examine the relationships between item responses after accounting for $\theta$ by performing residual correlation analysis [@chenLocalDependenceIndexes1997].

@@

In this paper, we use the most recent version of the RWA scale [@altemeyerAuthoritarians2006], which consists of 22 items in which participants are asked to indicate their agreement with the items on a nine-point scale ranging from "*strongly disagree*" (-4) to "*strongly agree*" (+4). While the RWA scale consists of three sub-dimensions, i.e., Submissiveness, Aggression, and Conventionalism, the RWA scale is theoretically assumed to be unidimensional because [@altemeyerAuthoritarians2006] shows that these sub-dimensions are strongly intercorrelated. While there is some evidence to the contrary [@duckittMultidimensionalityRightWingAuthoritarian2013], for the purposes of this article we will assume that the RWA scale is unidimensional.

@@

To maximize the reproducibility of our illustrative example, we wrote the article as a Quarto (.qmd) document, where we integrate the R codes used in the analysis as well as its outputs. We also include a longer, annotated R script file (.R) for the example we used as a supplementary document, which we highly recommend to use for a didactic purpose. The Quarto file (and its corresponding .docx and .pdf output) and R script file are publicly available on [a Github repository](). The dataset we used in this tutorial paper was obtained from a publicly accessible [Open-Source Psychometrics Project](http://openpsychometrics.org/_rawdata/).

@@

This command summarizes key descriptive statistics of a data frame, and the summary is provided in {apatb-table2}. As we see in {apatb-table2}, all variables have a minimum value of `0`, while the minimum score should be `1`.

Again, there is no information provided in the codebook regarding what `0` actually represents. Therefore, before deciding what we should do next, it is better to find out how many `0` we have in the data frame. To this end, we can calculate the percentage of `0` in each item by running these commands:

@@

After executing these commands, readers can see the percentage of `0` in each items in their R console, which overall are very small, ranging from 0.09% to 0.32% of the total cases. Since the percentages are very small, `0` is likely a code for a missing value. It is important to note that *mirt* package uses the Expected-Maximization (EM) algorithm to estimate model parameters, and the EM algorithm can handle missing values [@chalmersMirtMultidimensionalItem2012]. If we decide to eliminate cases with missing values, we still have a sizable number of cases, nonetheless. Therefore, to simplify our analysis, we only keep cases with complete responses by executing these commands:

@@

Interestingly, if we look at {apatb-table2} again, the mean scores of the items seem to differ drastically. Some items have high mean scores, ranging from 6 to 7, but some of those are comparatively low. The RWA scale indeed has some unfavorable items but, again, there is no information in the codebook whether these unfavorable items have been reversely coded. However, if we look closely at ten items with high mean ($\bar{x}$ \> 6), these are the items that are supposed to be unfavorable. Therefore, we can reasonably assume that unfavorable items have not been reversed in the original dataset, and we should do so by running these commands:

@@

In this step, we start to specify a vector representing our model, such as:

@@
